Spatial and temporal decay of SARS-CoV-2 environmental contamination of the healthcare environment

### 2021-09-16 manuscript.Rmd
Manuscript and Bib File Additions

* __Background__:

* __Methods__:

* __Results__:

* __Conclusions__:

@@

SARS-CoV-2 transmission can be mediated by droplets, aerosols, or fomites. Prior studies have demonstrated that SARS-CoV-2 extensively contaminates surfaces in the healthcare environment [@Morawska2021-mv]. But the spatial extent of contamination -- and how contamination changes during the course of hospitalization -- remain less well understood [@Mody2021-ag].

@@

* __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA. (IRB protocol #843273). Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability. All rooms underwent daily cleaning, according to hospital protocol. Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch objects (one wipe per surface). Floors were cleaned using a microfiber mop and a chlorine-based disinfectant (sodium dichloroisocyanurate).

@@

* __Specimen Collection and SARS-CoV-2 RT-PCR__: COVID-19 patient rooms were swabbed longitudinally until day 28 or patient discharge from the hospital (i.e., hospital days 1,2,7,14,21,and 28). Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room. Multiple surfaces were swabbed within each patient room using a Copan FLOQ(TM) swab and a sterile template (20cm2). Specific surfaces are listed in Table 1. Swabs were immediately placed into buffered AVL and frozen prior to further processing. Specimens then underwent RNA extraction and RT-PCR targeting the SARS-CoV-2 N1 region in triplicate.

@@

* __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces. There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of MDRO organisms [site ICEMAN ICHE paper], but the suface type was significantly associated with the probability of detection. Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% round(.,2)`) of SARS-CoV-2 detection, relative to elevated surfaces. High-touch elevated surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% round(.,2)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces. **Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.

@@

We sought to define the impact of time and distance on SARS-CoV-2 surface contamination through longitudinal sampling of surfaces within COVID-19 patient rooms. We found that distance from the patient's head of bed did not have a significant impact on the probability of detection of SARS-CoV-2 RNA. We had hypothesized that the probability of RNA detection with decrease with distance. However, we found that detection did not vary significantly among any surface group with increased distance. This is likely related to two factors 1) the relatively small size of patient rooms and 2) the physical spread of viral RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and HCP feet as well as the physical action of mopping which spread body fluids containing viral RNA within the patient's room. In contrast, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets on vertical surfaces. This finding matches previous description of high probability of floor contamination relative to other surfaces [@Chia2020].

While we did not observe a measurable impact of distance on the probability of surface contamination with SARS-CoV-2 RNA, we did detect a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis. A similar finding was described by Chia et al [@Chia2020], where patients over 7 days from illness had lower probability of surface contamination within rooms. Interestingly, while the probability of floor contamination decreased with time from diagnosis, the probability remained relatively high at later time points. Combined with the finding that the probability of floor contamination increased with time from the onset of the second COVID-19 Wave, this suggests that external contamination accounts for a significant proportion of floor contamination. Viral RNA carried on the feet of healthcare personnel or the mops of EVS teams may have resulted in increased probability of detection as COVID-19 census increased. Guo et al [@Guo2020] described that floor contamination was discovered in areas where clinical care was not performed, and suggested that it was spread by the feet of HCP.

Our findings are significant as they reveal that varied distances within a patient room have similar probability of surface contamination. This finding supports CDC recommended infection control practices [@CDCInfectionControl]. Additionally, our findings highlight the role of touch in surface contamination of patient rooms and the likelihood of extrinsic contamination on SARS-CoV-2 units.

Our study has several limitations. First, our study used quantitative PCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces. Secondly, we did not control for cleaning methods on study units. However, cleaning practices did not undergo significant changes during the study time period and sampling was performed routinely prior to daily cleaning. Finally, while we explored patient characteristics impacted the decay of surface contamination with time and distance, there was limited variability of some measures, making independent associations with viral detection less feasible.

In conclusion,   .

>>>>

* __Background__: SARS-CoV-2 RNA has been described to contaminate surfaces within healthcare settings. However, little is known regarding the role of distance and time in surface contamination in a clinical setting.

* __Methods__: We performed longitudinal sampling of surfaces within COVID-19 patient rooms while measuring the distance from each surface to the patient bed. We performed quantitative RT-PCR for the detection of SARS-CoV-2 RNA and modeled the role of distance, time, and severity of illness on the probability of detecting SARS-CoV-2 using a mixed-effects binomial model.

* __Results__: Probability of detecting SARS-CoV-2 RNA did not vary with distance. However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors: odds ratio [OR] 67.77 (95% confidence interval [CI] 36.34 to 131.47, high-touch elevated: OR 7.39 (95% CI 4.31 to 13.09)). Patients requiring high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation respiratory support had greater odds of  surface contamination (OR 1.76 (95% CrI 1.2 to 2.57)). While the probability of surface contamination decayed with prolonged hospitalization, probability of floor detection increased with time over the pandemic.

* __Conclusions__: SARS-CoV-2 surface deposition did not vary with distance within patient rooms. However, surface type, severity of illness, and time predicted surface deposition.

@@

SARS-CoV-2 transmission can be mediated by droplets, aerosols, or fomites. Prior studies have demonstrated that SARS-CoV-2 extensively contaminates surfaces in the healthcare environment [@Morawska2021-mv]. However, the spatial extent of contamination -- and how contamination changes during the course of hospitalization -- remain less well understood [@Mody2021-ag].

@@

* __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA. (IRB protocol #843273). Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability. All rooms underwent daily cleaning, according to hospital protocol. Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch objects, limiting use to a single surface. Floors were cleaned using a microfiber mop and a chlorine-based disinfectant (sodium dichloroisocyanurate).

@@

* __Specimen Collection and SARS-CoV-2 RT-PCR__: COVID-19 patient rooms were swabbed longitudinally until day 28 or patient discharge from the hospital (i.e., hospital days 1,2,7,14,21,and 28). Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room. Multiple surfaces were swabbed within each patient room using a Copan FLOQ(TM) swab and a sterile template (20cm2). Specific surfaces are listed in Table 1 with high-touch surfaces selected using structured observations prior to study start. Swabs were immediately placed into buffered AVL and frozen prior to further processing. Specimens then underwent RNA extraction and RT-PCR targeting the SARS-CoV-2 N1 region in triplicate.

@@

* __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces. There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of MDRO organisms [@Kelly2021], but the suface type was significantly associated with the probability of detection. Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% round(.,2)`) of SARS-CoV-2 detection, relative to elevated surfaces. High-touch elevated surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% round(.,2)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces. **Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.

@@

****ADD Discussion of our earlier ICHE paper

We sought to define the impact of time and distance on SARS-CoV-2 surface contamination through longitudinal sampling of  within COVID-19 patient rooms. We found that distance from the patient's bed did not inform the probability of detection of SARS-CoV-2 RNA. We had hypothesized that the probability of RNA detection would decrease with distance. This hypothesis was informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@KElly2021]. However, we found that SARS-CoV-2 detection did not vary significantly among any surface group as distance varied. This is most likely related to two factors 1) the relatively small size of patient rooms and 2) the physical spread of viral RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and healthcare personnel (HCP) foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient's room. This finding matches previous description of high probability of floor contamination relative to other surfaces [@Chia2020]. In contrast to floors, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.

While we did not observe a measurable impact of distance on the probability of surface contamination with SARS-CoV-2 RNA, we did detect a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis. A similar finding was described by Chia et al [@Chia2020], where patients over 7 days from illness had lower probability of surface contamination within rooms. Interestingly, while the probability of floor contamination decreased with time from diagnosis, the probability remained relatively high at later time points. Combined with the finding that the probability of floor contamination increased with time from the onset of the second COVID-19 Wave, this suggests that external contamination accounts for a proportion of floor contamination. Viral RNA carried on the feet of HCP or the mops of EVS teams may have resulted in increased probability of detection as COVID-19 census increased. This is supported by other research where floor contamination was discovered to spread beyond clinical areas [@Guo2020]. Additionally, it is possible that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.

With increased clinical severity, specifically those who required high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation, we detected an increased odds of detecting surface RNA within rooms. This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aeresol production associated with respiratory procedures. In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021] in a large observational study, potentially supporting the contribution of respiratory care in surface contamination. Thus, while, we can only hypothesize the mechanism for greater odds of detection in these patients, this finding supports CDC guidance for prolonged isolation in critically ill patients[@CDC2021].

Our study has several limitations. First, our study used quantitative PCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces. Secondly, we did not control for cleaning methods on study units. However, cleaning practices did not undergo systematic changes during the study time period and sampling was performed routinely prior to daily cleaning. Finally, we did not perform matching sampling of occupying patients to explore the role of patient viral load and surface contamination.

In conclusion, SARS-CoV-2 RNA surface deposition did not vary with distance within patient rooms. However, surface type, severity of illness, and time predicted detection of surface contamination. Future studies are needed to determine the role of surface contamination in risk of infection among patients and HCP.
